---
title: "Guides"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Mode selection

- `mode = "paper"`: adjacency = `r^2`.
- `mode = "python_compat"`: adjacency = `|r|`.

Use the same mode throughout a project to keep downstream interpretation stable.

## Parameter tuning

- `top_n_genes`: controls stage-1 feature space.
- `min_cgm_size`: controls coarse module granularity.
- `min_fgm_size`: controls fine module granularity.

Smaller minimum module sizes create more modules but increase fragmentation.

## Lightweight workflow pattern

```{r}
library(NestedWGCNA)
set.seed(200)
x <- matrix(rnorm(50 * 400), nrow = 50, ncol = 400)
colnames(x) <- paste0("G", seq_len(ncol(x)))

x2 <- top_variable_genes(x, n = 250)
cgm <- find_cgm(x2, mode = "python_compat", min_cgm_size = 40, keep_matrices = FALSE)
fgm <- find_fgm(x2, mode = "python_compat", min_fgm_size = 15)

head(cgm$assignment)
head(fgm$assignment)
```

This pattern avoids storing heavy intermediate matrices when `keep_matrices = FALSE`.
