---
title: "NestedWGCNA package article"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NestedWGCNA package article}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(NestedWGCNA)
read_summary <- function(dataset) {
  p <- file.path("inst", "extdata", "case_studies", dataset, "summary.json")
  if (!file.exists(p)) {
    p <- system.file("extdata", "case_studies", dataset, "summary.json", package = "NestedWGCNA")
  }
  if (file.exists(p)) jsonlite::fromJSON(p) else NULL
}
summary_tcga <- read_summary("tcga_blca")
summary_all <- read_summary("all_leukemia")
summary_bladder <- read_summary("bladderbatch")
```

# Overview

NestedWGCNA is an R package for two-stage gene co-expression analysis that separates broad biological programs from finer, within-program states.

Core workflow:

1. Build a gene-gene network and discover **coarse-grained modules (CGMs)**.
2. Extract CGM core genes and apply GenFocus normalization.
3. Re-cluster within CGMs to discover **fine-grained modules (FGMs)**.

Compared with classical WGCNA, NestedWGCNA uses UMAP + HDBSCAN and a nested decomposition strategy.

# Reproducibility policy

The package defines explicit method modes:

- `mode = "paper"`: follows the manuscript/README definition.
- `mode = "python_compat"`: reproduces current Python behavior.

This mode split is the main reproducibility contract.

# R implementation scope

This repository is R-only at runtime:

- all core methods are implemented in `R/`
- no Python runtime dependency is required
- case-study scripts are all `Rscript`-based

Executed real-data case-study scripts:

```text
Rscript inst/scripts/run_tcga_blca_case_study.R
Rscript inst/scripts/run_all_leukemia_case_study.R
Rscript inst/scripts/run_bladderbatch_case_study.R
```

# Real-data snapshot

The documentation is backed by executed analyses on three real datasets.

```{r}
summaries <- list(summary_tcga, summary_all, summary_bladder)
summaries <- summaries[!vapply(summaries, is.null, logical(1))]
if (length(summaries)) {
  tbl <- do.call(
    rbind,
    lapply(
      summaries,
      function(x) data.frame(
        dataset = x$dataset_id,
        samples = x$n_samples,
        input_genes = x$n_genes_input,
        modeled_genes = x$n_genes_model,
        target_cgm = x$target_cgm,
        cgm_modules = x$cgm_n_modules,
        fgm_modules = x$fgm_n_modules,
        stringsAsFactors = FALSE
      )
    )
  )
  knitr::kable(tbl, caption = "Executed real-data datasets")
}
```

# Output access API

`NestedWGCNA` provides accessors to bundled outputs:

```{r}
available_case_studies()
```

```{r}
case_study_summary("tcga_blca")$dataset_id
```

```{r}
head(case_study_table("all_leukemia", "phenotype_associations"))
```

# Network visualization API

Following WGCNA eigengene-network visualization conventions, the package now
provides module-network plotting helpers.

```{r}
tcga_scores <- case_study_table("tcga_blca", "cgm_scores")
rownames(tcga_scores) <- tcga_scores[[1]]
tcga_scores <- tcga_scores[, -1, drop = FALSE]

tcga_cgm_assign <- case_study_table("tcga_blca", "cgm_assignments")
tab <- table(tcga_cgm_assign$cgm[tcga_cgm_assign$cgm >= 0])
cgm_sizes <- stats::setNames(as.numeric(tab), paste0("CGM_", names(tab)))

op <- graphics::par(mfrow = c(1, 2), mar = c(1.5, 1.5, 2.8, 1))
plot_module_network(tcga_scores, module_sizes = cgm_sizes, layer = "CGM", top_n = 12, cor_thr = 0.25)
plot_module_score_heatmap(tcga_scores, layer = "CGM", top_n = 12)
graphics::par(op)
```

# Interpretation

Across datasets, CGMs capture high-level programs and FGMs resolve finer substructure within selected CGMs. This is the intended behavior of the nested design and is visible in all three tutorials.
