---
title: "NestedWGCNA package article"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NestedWGCNA package article}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
read_summary <- function(dataset) {
  p <- system.file("extdata", "case_studies", dataset, "summary.json", package = "NestedWGCNA")
  if (p == "") {
    p <- file.path("inst", "extdata", "case_studies", dataset, "summary.json")
  }
  if (file.exists(p)) jsonlite::fromJSON(p) else NULL
}
summary_tcga <- read_summary("tcga_blca")
summary_all <- read_summary("all_leukemia")
summary_bladder <- read_summary("bladderbatch")
```

# Why NestedWGCNA?

NestedWGCNA is a two-stage co-expression framework designed to separate broad biological programs from finer, within-program substructure.

The workflow is:

1. Build a gene-gene network and discover **coarse-grained modules (CGMs)**.
2. Extract CGM core genes and apply GenFocus normalization.
3. Re-cluster within CGMs to discover **fine-grained modules (FGMs)**.

Compared with classical WGCNA, the reference implementation uses UMAP + HDBSCAN and a nested decomposition strategy.

# Design policy for the R package

The package is being formalized around explicit method modes:

- `mode = "paper"`: follows the manuscript/README definition.
- `mode = "python_compat"`: reproduces current Python behavior.
- `mode = "custom"`: expert override for similarity/dissimilarity settings.

This mode split is the central reproducibility contract of the R package.

# Architecture

Repository layout is organized around:

- Core algorithm references (`inst/python_ref/utils.py`, `inst/python_ref/GenFocus.py`).
- R package scaffold (`DESCRIPTION`, `R/`, `tests/`, `vignettes/`).
- Reproducible real-data outputs (`inst/extdata/case_studies/`).

Case-study scripts are:

```text
python inst/scripts/run_tcga_blca_case_study.py
Rscript inst/scripts/run_all_leukemia_case_study.R
Rscript inst/scripts/run_bladderbatch_case_study.R
```

# Real-data snapshot (TCGA-BLCA)

The package docs are backed by an executed analysis on downloaded TCGA-BLCA data.

```{r}
if (!is.null(summary_tcga)) {
  key <- data.frame(
    metric = c(
      "samples",
      "input genes",
      "modeled genes",
      "CGM modules",
      "CGM assigned genes",
      "target CGM",
      "target CGM size",
      "INGS genes",
      "FGM modules",
      "FGM assigned genes"
    ),
    value = c(
      summary_tcga$n_samples,
      summary_tcga$n_genes_input,
      summary_tcga$n_genes_model,
      summary_tcga$cgm_n_modules,
      summary_tcga$cgm_assigned_genes,
      summary_tcga$target_cgm,
      summary_tcga$target_cgm_size,
      summary_tcga$genfocus_ings_n,
      summary_tcga$fgm_n_modules,
      summary_tcga$fgm_assigned_genes
    )
  )
  knitr::kable(key, caption = "TCGA-BLCA case-study summary")
}
```

```{r, echo = FALSE, fig.cap = "CGM module size distribution"}
img_cgm <- system.file(
  "extdata", "case_studies", "tcga_blca", "cgm_module_sizes.png", package = "NestedWGCNA"
)
if (img_cgm == "") {
  img_cgm <- file.path("inst", "extdata", "case_studies", "tcga_blca", "cgm_module_sizes.png")
}
if (file.exists(img_cgm)) {
  knitr::include_graphics(img_cgm)
}
```

```{r, echo = FALSE, fig.cap = "FGM module size distribution in the selected immune-like CGM"}
img_fgm <- system.file(
  "extdata", "case_studies", "tcga_blca", "fgm_module_sizes.png", package = "NestedWGCNA"
)
if (img_fgm == "") {
  img_fgm <- file.path("inst", "extdata", "case_studies", "tcga_blca", "fgm_module_sizes.png")
}
if (file.exists(img_fgm)) {
  knitr::include_graphics(img_fgm)
}
```

# Multi-dataset tutorial set

```{r}
summaries <- list(summary_tcga, summary_all, summary_bladder)
summaries <- summaries[!vapply(summaries, is.null, logical(1))]
if (length(summaries)) {
  tbl <- do.call(
    rbind,
    lapply(
      summaries,
      function(x) data.frame(
        dataset = x$dataset_id,
        samples = x$n_samples,
        input_genes = x$n_genes_input,
        modeled_genes = x$n_genes_model,
        cgm_modules = x$cgm_n_modules,
        fgm_modules = x$fgm_n_modules,
        stringsAsFactors = FALSE
      )
    )
  )
  knitr::kable(tbl, caption = "Executed real-data tutorial datasets")
}
```

# Interpretation in one paragraph

In this run, the CGM layer recovers broad stromal-like, epithelial-like, and macrophage-like structures, and the nested FGM layer further separates the immune-like CGM into an interferon/MHC-II branch and a cytotoxic T-cell branch. This is the expected qualitative behavior of the method: first capture coarse tissue programs, then resolve finer immune states after core-based normalization.
