---
title: "Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Function map

### Input and cleaning

- `as_expression_matrix()`
- `clear_data()`
- `top_variable_genes()`

### Similarity and dissimilarity

- `compute_adjacency()`
- `compute_dissimilarity()`

### Two-stage module inference

- `find_cgm()`
- `find_fgm()`
- `run_nested_wgcna()`

### Downstream compact outputs

- `module_score()`
- `summary()` for `NestedWGCNAResult`

## Result schema

```{r}
library(NestedWGCNA)
set.seed(400)
x <- matrix(rnorm(40 * 200), nrow = 40, ncol = 200)
colnames(x) <- paste0("G", seq_len(ncol(x)))
res <- run_nested_wgcna(x, min_cgm_size = 30, min_fgm_size = 10, top_n_genes = 150)
names(res)
```

## Notes

This package intentionally keeps the public API short and avoids optional heavy
subsystems to maintain a small footprint.
