---
title: "Tutorial: ALL leukemia real-data analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: ALL leukemia real-data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

locate_case_file <- function(name) {
  p <- system.file("extdata", "case_studies", "all_leukemia", name, package = "NestedWGCNA")
  if (p == "") {
    p <- file.path("inst", "extdata", "case_studies", "all_leukemia", name)
  }
  p
}

summary_path <- locate_case_file("summary.json")
assoc_path <- locate_case_file("phenotype_associations.tsv")

summary_obj <- if (file.exists(summary_path)) jsonlite::fromJSON(summary_path) else NULL
assoc <- if (file.exists(assoc_path)) read.delim(assoc_path, check.names = FALSE) else NULL
```

# Background

The Bioconductor `ALL` dataset is a canonical acute lymphoblastic leukemia expression cohort with immunophenotype (`BT`) and molecular subtype labels (`mol.biol`). It is a strong test bed for whether nested co-expression modules capture clinically meaningful leukemia structure.

# Objective

The goal is to test if NestedWGCNA modules are associated with known ALL subtype labels, beyond global gene-expression trends.

# Data source and execution

This tutorial uses real data downloaded through Bioconductor package installation:

- Dataset: `ALL::ALL` ExpressionSet
- Samples: 128
- Features (probes): 12,625

Run:

```{r, eval = FALSE}
system("Rscript inst/scripts/run_all_leukemia_case_study.R")
```

# Methods

Pipeline settings in this run:

1. Select top 1,200 variable probes.
2. CGM discovery with `min_cluster_size = 30`.
3. GenFocus inside the selected CGM.
4. FGM discovery with requested `min_cluster_size = 8`.
5. Association testing of module scores against `BT`, `mol.biol`, `sex`, and `age`.

```{r}
if (!is.null(summary_obj)) {
  method_tbl <- data.frame(
    item = c(
      "samples",
      "input probes",
      "modeled probes",
      "CGM modules",
      "FGM modules",
      "GenFocus corr_thr",
      "GenFocus INGS"
    ),
    value = c(
      summary_obj$n_samples,
      summary_obj$n_genes_input,
      summary_obj$n_genes_model,
      summary_obj$cgm_n_modules,
      summary_obj$fgm_n_modules,
      summary_obj$genfocus_corr_thr,
      summary_obj$genfocus_ings_n
    )
  )
  knitr::kable(method_tbl, caption = "ALL case-study summary")
}
```

# Results

```{r}
if (!is.null(summary_obj)) {
  res_tbl <- data.frame(
    layer = c("CGM", "FGM"),
    modules = c(summary_obj$cgm_n_modules, summary_obj$fgm_n_modules),
    assigned_genes = c(summary_obj$cgm_assigned_genes, summary_obj$fgm_assigned_genes),
    noise_genes = c(summary_obj$cgm_noise_genes, summary_obj$fgm_noise_genes)
  )
  knitr::kable(res_tbl, caption = "Recovered module structure")
}
```

```{r, echo = FALSE, fig.cap = "CGM module size distribution (ALL)"}
img <- locate_case_file("cgm_module_sizes.png")
if (file.exists(img)) knitr::include_graphics(img)
```

```{r, echo = FALSE, fig.cap = "FGM module size distribution (ALL)"}
img <- locate_case_file("fgm_module_sizes.png")
if (file.exists(img)) knitr::include_graphics(img)
```

```{r}
if (!is.null(assoc) && nrow(assoc) > 0) {
  top_assoc <- assoc[order(assoc$fdr), c("layer", "module", "phenotype", "test", "effect", "fdr")]
  top_assoc <- head(top_assoc, 12)
  knitr::kable(top_assoc, digits = 4, caption = "Top phenotype associations")
}
```

Main observation from this run:

- Strong FGM associations with `BT` (immunophenotype) and `mol.biol` (molecular subtype).
- The strongest association was FGM-level, indicating Stage 2 refinement carries subtype-discriminating information.

# Discussion

In ALL, subtype signal is expected but can be distributed across multiple pathways. The nested design helps by first collecting broad leukemia co-expression structure, then splitting it into finer modules that align more directly with subtype labels.

# Interpretation

For downstream use, CGMs provide a stable high-level representation, while FGMs capture subtype-sensitive variation suitable for stratification tasks. In this run, FGM-level associations with `BT`/`mol.biol` were stronger than coarse-level associations, consistent with the intended purpose of the two-stage method.
