---
title: "Tutorial: ALL leukemia real-data analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: ALL leukemia real-data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(NestedWGCNA)

locate_case_file <- function(name) {
  p <- file.path("inst", "extdata", "case_studies", "all_leukemia", name)
  if (!file.exists(p)) {
    p <- system.file("extdata", "case_studies", "all_leukemia", name, package = "NestedWGCNA")
  }
  p
}

summary_path <- locate_case_file("summary.json")
assoc_path <- locate_case_file("phenotype_associations.tsv")
cgm_enr_path <- locate_case_file("cgm_enrichment.tsv")
fgm_enr_path <- locate_case_file("fgm_enrichment.tsv")
cgm_assign_path <- locate_case_file("cgm_assignments.tsv")
fgm_assign_path <- locate_case_file("fgm_assignments.tsv")
cgm_scores_path <- locate_case_file("cgm_scores.tsv")
fgm_scores_path <- locate_case_file("fgm_scores.tsv")

read_case_tsv <- function(path, row_names = FALSE) {
  if (!file.exists(path)) {
    return(NULL)
  }
  sz <- file.info(path)$size
  if (!is.finite(sz) || sz <= 0) {
    return(NULL)
  }
  tryCatch(
    if (row_names) {
      read.delim(path, check.names = FALSE, row.names = 1)
    } else {
      read.delim(path, check.names = FALSE)
    },
    error = function(e) NULL
  )
}

is_empty_tbl <- function(x) is.null(x) || !nrow(x)

summary_obj <- if (file.exists(summary_path)) jsonlite::fromJSON(summary_path) else NULL
assoc <- read_case_tsv(assoc_path)
cgm_enr <- read_case_tsv(cgm_enr_path)
fgm_enr <- read_case_tsv(fgm_enr_path)
cgm_assign <- read_case_tsv(cgm_assign_path)
fgm_assign <- read_case_tsv(fgm_assign_path)
cgm_scores <- read_case_tsv(cgm_scores_path, row_names = TRUE)
fgm_scores <- read_case_tsv(fgm_scores_path, row_names = TRUE)
val_or_na <- function(x) if (is.null(x)) NA else x

module_size_map <- function(assign_df, module_col, prefix) {
  if (is.null(assign_df) || !nrow(assign_df) || !module_col %in% colnames(assign_df)) {
    return(NULL)
  }
  ids <- suppressWarnings(as.integer(assign_df[[module_col]]))
  ids <- ids[is.finite(ids) & ids >= 0]
  if (!length(ids)) {
    return(NULL)
  }
  tab <- table(ids)
  stats::setNames(as.numeric(tab), paste0(prefix, names(tab)))
}

cgm_sizes <- module_size_map(cgm_assign, "cgm", "CGM_")
fgm_sizes <- module_size_map(fgm_assign, "fgm", "FGM_")
```

# Background

The Bioconductor `ALL` dataset is a canonical acute lymphoblastic leukemia expression cohort with immunophenotype (`BT`) and molecular subtype labels (`mol.biol`). It is a strong test bed for whether nested co-expression modules capture clinically meaningful leukemia structure.

# Objective

The goal is to test if NestedWGCNA modules are associated with known ALL subtype labels, beyond global gene-expression trends.

# Data source and execution

This tutorial uses real data downloaded through Bioconductor package installation:

- Dataset: `ALL::ALL` ExpressionSet
- Samples: 128
- Features (probes): 12,625

Run:

```{r, eval = FALSE}
system("Rscript inst/scripts/run_all_leukemia_case_study.R")
```

# Methods

Pipeline settings in this run:

1. Select top 1,200 variable probes.
2. CGM discovery with `min_cluster_size = 30`.
3. Attempt GenFocus inside the selected CGM; if INGS is not established, use direct FGM clustering on the CGM matrix.
4. FGM discovery with requested `min_cluster_size = 8`.
5. Association testing of module scores against `BT`, `mol.biol`, `sex`, and `age`.

```{r}
if (!is.null(summary_obj)) {
  method_tbl <- data.frame(
    item = c(
      "samples",
      "input probes",
      "modeled probes",
      "CGM modules",
      "FGM modules",
      "GenFocus corr_thr",
      "GenFocus INGS"
    ),
    value = c(
      summary_obj$n_samples,
      summary_obj$n_genes_input,
      summary_obj$n_genes_model,
      summary_obj$cgm_n_modules,
      summary_obj$fgm_n_modules,
      val_or_na(summary_obj$genfocus_corr_thr),
      summary_obj$genfocus_ings_n
    )
  )
  knitr::kable(method_tbl, caption = "ALL case-study summary")
}
```

# Results

```{r}
if (!is.null(summary_obj)) {
  res_tbl <- data.frame(
    layer = c("CGM", "FGM"),
    modules = c(summary_obj$cgm_n_modules, summary_obj$fgm_n_modules),
    assigned_genes = c(summary_obj$cgm_assigned_genes, summary_obj$fgm_assigned_genes),
    noise_genes = c(summary_obj$cgm_noise_genes, summary_obj$fgm_noise_genes)
  )
  knitr::kable(res_tbl, caption = "Recovered module structure")
}
```

```{r, echo = FALSE, fig.cap = "CGM module size distribution (ALL)"}
img <- "figures/all_leukemia_cgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

```{r, echo = FALSE, fig.cap = "FGM module size distribution (ALL)"}
img <- "figures/all_leukemia_fgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

```{r}
if (!is.null(assoc) && nrow(assoc) > 0) {
  top_assoc <- assoc[order(assoc$fdr), c("layer", "module", "phenotype", "test", "effect", "fdr")]
  top_assoc <- head(top_assoc, 12)
  knitr::kable(top_assoc, digits = 4, caption = "Top phenotype associations")
}
```

## Network diagram

```{r, fig.cap = "Force-directed module networks in ALL (CGM and FGM layers)"}
op <- graphics::par(mfrow = c(1, 2), mar = c(1.5, 1.5, 2.8, 1))
plot_module_network(cgm_scores, module_sizes = cgm_sizes, layer = "CGM", top_n = 12, cor_thr = 0.25)
plot_module_network(fgm_scores, module_sizes = fgm_sizes, layer = "FGM", top_n = 12, cor_thr = 0.25)
graphics::par(op)
```

```{r, fig.cap = "WGCNA-style module-score correlation heatmap (CGM layer)"}
plot_module_score_heatmap(cgm_scores, layer = "CGM", top_n = 15)
```

```{r, fig.cap = "WGCNA-style module-score correlation heatmap (FGM layer)"}
plot_module_score_heatmap(fgm_scores, layer = "FGM", top_n = 15)
```

## Quantitative readout and objective match

```{r}
if (!is.null(cgm_assign)) {
  size_tbl <- sort(table(cgm_assign$cgm[cgm_assign$cgm >= 0]), decreasing = TRUE)
  size_tbl <- data.frame(cgm = names(size_tbl), genes = as.integer(size_tbl))
  knitr::kable(head(size_tbl, 7), caption = "Largest CGMs by gene count")
}
```

```{r}
if (!is.null(fgm_assign)) {
  size_tbl <- sort(table(fgm_assign$fgm[fgm_assign$fgm >= 0]), decreasing = TRUE)
  size_tbl <- data.frame(fgm = names(size_tbl), genes = as.integer(size_tbl))
  knitr::kable(head(size_tbl, 7), caption = "Largest FGMs by gene count")
}
```

```{r}
if (!is.null(assoc) && nrow(assoc) > 0) {
  layer_summary <- lapply(split(assoc, assoc$layer), function(z) {
    data.frame(
      layer = unique(z$layer),
      tests = nrow(z),
      significant_fdr_0_05 = sum(z$fdr < 0.05),
      best_fdr = min(z$fdr),
      stringsAsFactors = FALSE
    )
  })
  layer_summary <- do.call(rbind, layer_summary)
  knitr::kable(layer_summary, digits = 4, caption = "Association strength by layer")
}
```

```{r}
if (!is.null(assoc) && nrow(assoc) > 0) {
  fgm_assoc <- assoc[assoc$layer == "FGM", c("layer", "module", "phenotype", "test", "effect", "fdr")]
  fgm_assoc <- fgm_assoc[order(fgm_assoc$fdr), , drop = FALSE]
  if (nrow(fgm_assoc)) {
    knitr::kable(head(fgm_assoc, 8), digits = 4, caption = "Top FGM phenotype associations")
  }
}
```

```{r}
if (is_empty_tbl(cgm_enr) && is_empty_tbl(fgm_enr)) {
  knitr::kable(
    data.frame(
      note = "No xCell/BostonGene enrichment was detected in this run, likely because probe IDs are not directly mapped to gene symbols in the bundled matrix."
    ),
    caption = "Enrichment interpretation note"
  )
}
```

Main observation from this run:

- Strongest statistical signals are CGM-level and map to known ALL labels (`BT`, `mol.biol`), with best FDR around `1e-13`.
- FGMs still add subtype-linked signal (best FGM FDR around `1e-4`), but at lower magnitude than coarse modules.
- Network topology is layered: CGM graph captures broad leukemia structure, and FGM graph shows finer within-target relationships.

# Discussion

In this microarray ALL cohort, known subtype structure is dominant, so it is expected that CGM-level scores carry most of the effect size. The results match that expectation: many CGM tests remain significant after FDR correction, while FGM signals are fewer and weaker but still detectable.

The network diagram supports this reading. CGM modules form a clear global backbone corresponding to broad subtype organization. FGM modules contribute local edges and secondary partitions, indicating that nested decomposition is not redundant even when coarse subtype effects are strong.

Biological enrichment is empty in this run because the packaged matrix uses probe identifiers and the included signature collections use gene symbols. This is an annotation-compatibility issue, not a failure of the module-discovery step.

# Interpretation

For practical ALL modeling, use CGM scores as the primary covariates for robust subtype-scale structure, then add selected FGMs when finer separation is needed for subgrouping. In this dataset, the objective is achieved: stage 1 recovers the dominant leukemia axes, and stage 2 adds interpretable substructure without replacing the coarse signal.

# Visualization design note

The network and heatmap style in this tutorial is aligned with:

- NestedWGCNA project structure and reporting style: <https://github.com/ilyada/NestedWGCNA>
- WGCNA eigengene-network visualization (`plotEigengeneNetworks`): <https://rdrr.io/cran/WGCNA/man/plotEigengeneNetworks.html>
- WGCNA network export practice (`exportNetworkToCytoscape`): <https://rdrr.io/cran/WGCNA/man/exportNetworkToCytoscape.html>
