---
title: "Tutorial: bladderbatch real-data analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: bladderbatch real-data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(NestedWGCNA)

locate_case_file <- function(name) {
  p <- file.path("inst", "extdata", "case_studies", "bladderbatch", name)
  if (!file.exists(p)) {
    p <- system.file("extdata", "case_studies", "bladderbatch", name, package = "NestedWGCNA")
  }
  p
}

summary_path <- locate_case_file("summary.json")
assoc_path <- locate_case_file("phenotype_associations.tsv")
cgm_enr_path <- locate_case_file("cgm_enrichment.tsv")
fgm_enr_path <- locate_case_file("fgm_enrichment.tsv")
cgm_assign_path <- locate_case_file("cgm_assignments.tsv")
fgm_assign_path <- locate_case_file("fgm_assignments.tsv")
cgm_scores_path <- locate_case_file("cgm_scores.tsv")
fgm_scores_path <- locate_case_file("fgm_scores.tsv")

read_case_tsv <- function(path, row_names = FALSE) {
  if (!file.exists(path)) {
    return(NULL)
  }
  sz <- file.info(path)$size
  if (!is.finite(sz) || sz <= 0) {
    return(NULL)
  }
  tryCatch(
    if (row_names) {
      read.delim(path, check.names = FALSE, row.names = 1)
    } else {
      read.delim(path, check.names = FALSE)
    },
    error = function(e) NULL
  )
}

is_empty_tbl <- function(x) is.null(x) || !nrow(x)

summary_obj <- if (file.exists(summary_path)) jsonlite::fromJSON(summary_path) else NULL
assoc <- read_case_tsv(assoc_path)
cgm_enr <- read_case_tsv(cgm_enr_path)
fgm_enr <- read_case_tsv(fgm_enr_path)
cgm_assign <- read_case_tsv(cgm_assign_path)
fgm_assign <- read_case_tsv(fgm_assign_path)
cgm_scores <- read_case_tsv(cgm_scores_path, row_names = TRUE)
fgm_scores <- read_case_tsv(fgm_scores_path, row_names = TRUE)

module_size_map <- function(assign_df, module_col, prefix) {
  if (is.null(assign_df) || !nrow(assign_df) || !module_col %in% colnames(assign_df)) {
    return(NULL)
  }
  ids <- suppressWarnings(as.integer(assign_df[[module_col]]))
  ids <- ids[is.finite(ids) & ids >= 0]
  if (!length(ids)) {
    return(NULL)
  }
  tab <- table(ids)
  stats::setNames(as.numeric(tab), paste0(prefix, names(tab)))
}

cgm_sizes <- module_size_map(cgm_assign, "cgm", "CGM_")
fgm_sizes <- module_size_map(fgm_assign, "fgm", "FGM_")
```

# Background

The Bioconductor `bladderbatch` dataset contains bladder-related expression profiles with sample labels for cancer status (`cancer`) and study outcome group (`outcome`), plus batch labels. It is suitable for evaluating whether NestedWGCNA detects biologically and clinically structured modules in a small-to-medium cohort.

# Objective

The goal is to test whether CGM/FGM module scores are associated with `cancer` and `outcome` labels in real bladder data.

# Data source and execution

This tutorial uses real data downloaded through Bioconductor package installation:

- Dataset: `bladderbatch::bladderEset`
- Samples: 57
- Features (probes): 22,283

Run:

```{r, eval = FALSE}
system("Rscript inst/scripts/run_bladderbatch_case_study.R")
```

# Methods

Pipeline settings in this run:

1. Select top 1,200 variable probes.
2. CGM discovery with `min_cluster_size = 12`.
3. Attempt GenFocus inside the target CGM; if INGS is not established, continue with direct FGM clustering.
4. FGM discovery with requested `min_cluster_size = 6`.
5. Module score association against `cancer`, `outcome`, and `batch`.

```{r}
if (!is.null(summary_obj)) {
  method_tbl <- data.frame(
    item = c(
      "samples",
      "input probes",
      "modeled probes",
      "CGM modules",
      "FGM modules",
      "target CGM",
      "GenFocus INGS"
    ),
    value = c(
      summary_obj$n_samples,
      summary_obj$n_genes_input,
      summary_obj$n_genes_model,
      summary_obj$cgm_n_modules,
      summary_obj$fgm_n_modules,
      summary_obj$target_cgm,
      summary_obj$genfocus_ings_n
    )
  )
  knitr::kable(method_tbl, caption = "bladderbatch case-study summary")
}
```

# Results

```{r}
if (!is.null(summary_obj)) {
  res_tbl <- data.frame(
    layer = c("CGM", "FGM"),
    modules = c(summary_obj$cgm_n_modules, summary_obj$fgm_n_modules),
    assigned_genes = c(summary_obj$cgm_assigned_genes, summary_obj$fgm_assigned_genes),
    noise_genes = c(summary_obj$cgm_noise_genes, summary_obj$fgm_noise_genes)
  )
  knitr::kable(res_tbl, caption = "Recovered module structure")
}
```

```{r, echo = FALSE, fig.cap = "CGM module size distribution (bladderbatch)"}
img <- "figures/bladderbatch_cgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

```{r, echo = FALSE, fig.cap = "FGM module size distribution (bladderbatch)"}
img <- "figures/bladderbatch_fgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

```{r}
if (!is.null(assoc) && nrow(assoc) > 0) {
  top_assoc <- assoc[order(assoc$fdr), c("layer", "module", "phenotype", "test", "effect", "fdr")]
  top_assoc <- head(top_assoc, 12)
  knitr::kable(top_assoc, digits = 4, caption = "Top phenotype associations")
}
```

## Network diagram

```{r, fig.cap = "Force-directed module networks in bladderbatch (CGM and FGM layers)"}
op <- graphics::par(mfrow = c(1, 2), mar = c(1.5, 1.5, 2.8, 1))
plot_module_network(cgm_scores, module_sizes = cgm_sizes, layer = "CGM", top_n = 12, cor_thr = 0.25)
plot_module_network(fgm_scores, module_sizes = fgm_sizes, layer = "FGM", top_n = 12, cor_thr = 0.25)
graphics::par(op)
```

```{r, fig.cap = "WGCNA-style module-score correlation heatmap (CGM layer)"}
plot_module_score_heatmap(cgm_scores, layer = "CGM", top_n = 15)
```

```{r, fig.cap = "WGCNA-style module-score correlation heatmap (FGM layer)"}
plot_module_score_heatmap(fgm_scores, layer = "FGM", top_n = 15)
```

## Quantitative readout and objective match

```{r}
if (!is.null(cgm_assign)) {
  size_tbl <- sort(table(cgm_assign$cgm[cgm_assign$cgm >= 0]), decreasing = TRUE)
  size_tbl <- data.frame(cgm = names(size_tbl), genes = as.integer(size_tbl))
  knitr::kable(head(size_tbl, 10), caption = "Largest CGMs by gene count")
}
```

```{r}
if (!is.null(fgm_assign)) {
  size_tbl <- sort(table(fgm_assign$fgm[fgm_assign$fgm >= 0]), decreasing = TRUE)
  size_tbl <- data.frame(fgm = names(size_tbl), genes = as.integer(size_tbl))
  knitr::kable(head(size_tbl, 10), caption = "Largest FGMs by gene count")
}
```

```{r}
if (!is.null(assoc) && nrow(assoc) > 0) {
  layer_summary <- lapply(split(assoc, assoc$layer), function(z) {
    data.frame(
      layer = unique(z$layer),
      tests = nrow(z),
      significant_fdr_0_05 = sum(z$fdr < 0.05),
      best_fdr = min(z$fdr),
      stringsAsFactors = FALSE
    )
  })
  layer_summary <- do.call(rbind, layer_summary)
  knitr::kable(layer_summary, digits = 4, caption = "Association strength by layer")
}
```

```{r}
if (!is.null(assoc) && nrow(assoc) > 0) {
  fgm_assoc <- assoc[assoc$layer == "FGM", c("layer", "module", "phenotype", "test", "effect", "fdr")]
  fgm_assoc <- fgm_assoc[order(fgm_assoc$fdr), , drop = FALSE]
  if (nrow(fgm_assoc)) {
    knitr::kable(head(fgm_assoc, 8), digits = 4, caption = "Top FGM phenotype associations")
  }
}
```

```{r}
if (is_empty_tbl(cgm_enr) && is_empty_tbl(fgm_enr)) {
  knitr::kable(
    data.frame(
      note = "No xCell/BostonGene enrichment was detected in this run, likely because probe IDs are not directly mapped to gene symbols in the bundled matrix."
    ),
    caption = "Enrichment interpretation note"
  )
}
```

Main observation from this run:

- CGM-level associations with `outcome` and `cancer` were strong and reproducible (best FDR around `1e-7`).
- FGM-level tests did not pass FDR `< 0.05` in this small cohort, indicating limited power for fine stratification at stage 2.
- The module-score network still shows structured dependencies, so FGMs can be treated as exploratory hypotheses rather than definitive phenotype markers.

# Discussion

This cohort is small (`n = 57`) and has many modeled probes, so variance in fine-level association testing is expected. The nested pipeline still provides value: coarse modules identify robust disease-relevant axes, while fine modules partition those axes for follow-up exploration.

The network diagram clarifies this behavior. CGM connectivity reveals stable global organization linked to phenotype labels, while FGM connectivity captures weaker but non-random local structure inside the selected target module. This is the expected pattern when stage 1 is high-signal and stage 2 is power-limited.

As with the ALL tutorial, signature enrichment is empty here because the expression features are probes rather than harmonized gene symbols for the included signature collections.

# Interpretation

For bladderbatch, CGMs should be interpreted as primary disease programs and used first for inferential analyses. FGMs are better treated as candidate subprograms for validation in larger cohorts or matched RNA-seq datasets. The objective is partially achieved: stage 1 yields strong phenotype coupling, stage 2 adds interpretable but currently underpowered fine structure.

# Visualization design note

The network and heatmap style in this tutorial is aligned with:

- NestedWGCNA project structure and reporting style: <https://github.com/ilyada/NestedWGCNA>
- WGCNA eigengene-network visualization (`plotEigengeneNetworks`): <https://rdrr.io/cran/WGCNA/man/plotEigengeneNetworks.html>
- WGCNA network export practice (`exportNetworkToCytoscape`): <https://rdrr.io/cran/WGCNA/man/exportNetworkToCytoscape.html>
