---
title: "Tutorial: TCGA-BLCA real-data analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: TCGA-BLCA real-data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(NestedWGCNA)

locate_case_file <- function(name) {
  p <- file.path("inst", "extdata", "case_studies", "tcga_blca", name)
  if (!file.exists(p)) {
    p <- system.file("extdata", "case_studies", "tcga_blca", name, package = "NestedWGCNA")
  }
  p
}

summary_path <- locate_case_file("summary.json")
cgm_enr_path <- locate_case_file("cgm_enrichment.tsv")
fgm_enr_path <- locate_case_file("fgm_enrichment.tsv")
cgm_assign_path <- locate_case_file("cgm_assignments.tsv")
fgm_assign_path <- locate_case_file("fgm_assignments.tsv")
cgm_scores_path <- locate_case_file("cgm_scores.tsv")
fgm_scores_path <- locate_case_file("fgm_scores.tsv")

read_case_tsv <- function(path, row_names = FALSE) {
  if (!file.exists(path)) {
    return(NULL)
  }
  sz <- file.info(path)$size
  if (!is.finite(sz) || sz <= 0) {
    return(NULL)
  }
  tryCatch(
    if (row_names) {
      read.delim(path, check.names = FALSE, row.names = 1)
    } else {
      read.delim(path, check.names = FALSE)
    },
    error = function(e) NULL
  )
}

summary_obj <- if (file.exists(summary_path)) jsonlite::fromJSON(summary_path) else NULL
cgm_enr <- read_case_tsv(cgm_enr_path)
fgm_enr <- read_case_tsv(fgm_enr_path)
cgm_assign <- read_case_tsv(cgm_assign_path)
fgm_assign <- read_case_tsv(fgm_assign_path)
cgm_scores <- read_case_tsv(cgm_scores_path, row_names = TRUE)
fgm_scores <- read_case_tsv(fgm_scores_path, row_names = TRUE)
val_or_na <- function(x) if (is.null(x)) NA else x

plot_module_corr_network <- function(score_df, layer = "CGM", top_n = 12, cor_thr = 0.35) {
  if (is.null(score_df) || !nrow(score_df) || ncol(score_df) < 2) {
    plot.new()
    graphics::text(0.5, 0.5, paste(layer, "network unavailable"))
    return(invisible(NULL))
  }

  s <- as.matrix(score_df)
  storage.mode(s) <- "double"
  vars <- apply(s, 2, stats::var, na.rm = TRUE)
  vars[!is.finite(vars)] <- 0
  keep <- order(vars, decreasing = TRUE)[seq_len(min(top_n, ncol(s)))]
  s <- s[, keep, drop = FALSE]
  node_names <- colnames(s)

  r <- suppressWarnings(stats::cor(s, method = "spearman", use = "pairwise.complete.obs"))
  r[!is.finite(r)] <- 0
  diag(r) <- 0

  edge_idx <- which(abs(r) >= cor_thr, arr.ind = TRUE)
  edge_idx <- edge_idx[edge_idx[, 1] < edge_idx[, 2], , drop = FALSE]
  if (!nrow(edge_idx)) {
    upp <- abs(r[upper.tri(r)])
    fallback <- if (length(upp)) max(0.2, as.numeric(stats::quantile(upp, probs = 0.85, na.rm = TRUE))) else 1
    edge_idx <- which(abs(r) >= fallback, arr.ind = TRUE)
    edge_idx <- edge_idx[edge_idx[, 1] < edge_idx[, 2], , drop = FALSE]
  }

  d <- stats::as.dist(1 - abs(r))
  coords <- try(stats::cmdscale(d, k = 2), silent = TRUE)
  if (inherits(coords, "try-error") || any(!is.finite(coords))) {
    th <- seq(0, 2 * pi, length.out = length(node_names) + 1)[-1]
    coords <- cbind(cos(th), sin(th))
  }
  rownames(coords) <- node_names
  colnames(coords) <- c("x", "y")

  xr <- range(coords[, "x"])
  yr <- range(coords[, "y"])
  if (diff(xr) > 0) {
    coords[, "x"] <- (coords[, "x"] - mean(xr)) / diff(xr) * 2
  }
  if (diff(yr) > 0) {
    coords[, "y"] <- (coords[, "y"] - mean(yr)) / diff(yr) * 2
  }

  graphics::plot(coords[, "x"], coords[, "y"], type = "n", axes = FALSE, xlab = "", ylab = "",
                 main = paste0(layer, " module-score network"))
  if (nrow(edge_idx)) {
    for (i in seq_len(nrow(edge_idx))) {
      a <- edge_idx[i, 1]
      b <- edge_idx[i, 2]
      w <- r[a, b]
      edge_col <- if (w >= 0) "#1f77b4AA" else "#d62728AA"
      graphics::segments(coords[a, "x"], coords[a, "y"], coords[b, "x"], coords[b, "y"],
                         col = edge_col, lwd = 1 + 4 * abs(w))
    }
  }
  graphics::points(coords[, "x"], coords[, "y"], pch = 21, cex = 1.8, bg = "#f4a261", col = "#264653")
  graphics::text(coords[, "x"], coords[, "y"], labels = node_names, pos = 3, cex = 0.72)
  graphics::legend("topright", legend = c("positive", "negative"), col = c("#1f77b4", "#d62728"),
                   lwd = 2, bty = "n", cex = 0.8)
  invisible(r)
}
```

# Background

Bulk RNA-seq co-expression analysis often mixes strong tissue-level signals with finer immune-state signals. A single module scale can miss this hierarchy. NestedWGCNA addresses this by first identifying broad modules (CGMs), then re-normalizing and re-clustering inside each module to reveal FGM structure.

# Objective

The objective of this tutorial is to run a fully reproducible, real-data NestedWGCNA workflow on TCGA-BLCA and interpret whether the two-stage design separates broad tumor programs from finer immune programs.

# Data source and download

This tutorial uses an openly available TCGA-BLCA expression matrix:

- URL: `https://raw.githubusercontent.com/ilyada/NestedWGCNA/main/data/TCGA_BLCA.tsv`
- Samples: 406
- Genes: 20,017

The analysis script downloads the expression matrix and signature references, runs the full R pipeline, and writes outputs under `inst/extdata/case_studies/tcga_blca/`.

```{r, eval = FALSE}
system("Rscript inst/scripts/run_tcga_blca_case_study.R")
```

# Methods

The executed case-study configuration is:

1. Log-transform expression and keep top 1,200 variable genes.
2. Run CGM discovery (`min_cluster_size = 50`) with `find_cgm()`.
3. Select an immune-like CGM by signature enrichment.
4. Attempt GenFocus (`focus = "eigengene"`, `corr_thr` ladder from 0.9 to 0.7, `CVR_thr = 0.6`).
5. If no INGS is established, run FGM discovery directly on the target CGM expression matrix.
6. Run FGM discovery in that CGM (`min_cluster_size = 10`).
7. Perform hypergeometric enrichment against xCell and BostonGene signatures.

```{r}
if (!is.null(summary_obj)) {
  method_tbl <- data.frame(
    item = c(
      "samples",
      "input genes",
      "modeled genes",
      "CGM min_cluster_size",
      "FGM min_cluster_size",
      "GenFocus corr_thr",
      "GenFocus INGS genes"
    ),
    value = c(
      summary_obj$n_samples,
      summary_obj$n_genes_input,
      summary_obj$n_genes_model,
      summary_obj$cgm_min_cluster_size,
      summary_obj$fgm_min_cluster_size_used,
      val_or_na(summary_obj$genfocus_corr_thr),
      summary_obj$genfocus_ings_n
    )
  )
  knitr::kable(method_tbl, caption = "Case-study analysis settings")
}
```

# Results

## Module recovery

```{r}
if (!is.null(summary_obj)) {
  res_tbl <- data.frame(
    layer = c("CGM", "FGM (inside target CGM)"),
    modules = c(summary_obj$cgm_n_modules, summary_obj$fgm_n_modules),
    assigned_genes = c(summary_obj$cgm_assigned_genes, summary_obj$fgm_assigned_genes),
    noise_genes = c(summary_obj$cgm_noise_genes, summary_obj$fgm_noise_genes)
  )
  knitr::kable(res_tbl, caption = "Recovered modules")
}
```

```{r, echo = FALSE, fig.cap = "CGM sizes"}
img <- "figures/tcga_blca_cgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

```{r, echo = FALSE, fig.cap = "FGM sizes in target CGM"}
img <- "figures/tcga_blca_fgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

## Network diagram

```{r, fig.cap = "Module-score correlation networks in TCGA-BLCA (CGM and FGM layers)"}
op <- graphics::par(mfrow = c(1, 2), mar = c(1.5, 1.5, 2.8, 1))
plot_module_corr_network(cgm_scores, layer = "CGM", top_n = 12, cor_thr = 0.3)
plot_module_corr_network(fgm_scores, layer = "FGM", top_n = 12, cor_thr = 0.3)
graphics::par(op)
```

## Biological enrichment

```{r}
if (!is.null(cgm_enr) && nrow(cgm_enr) > 0) {
  cgm_top <- cgm_enr[order(cgm_enr$fdr), c("module", "source", "signature", "k", "K", "fdr")]
  cgm_top <- head(cgm_top, 10)
  knitr::kable(cgm_top, digits = 4, caption = "Top CGM enrichments")
}
```

```{r}
if (!is.null(fgm_enr) && nrow(fgm_enr) > 0) {
  fgm_top <- fgm_enr[order(fgm_enr$fdr), c("fgm", "source", "signature", "k", "K", "fdr")]
  fgm_top <- head(fgm_top, 10)
  knitr::kable(fgm_top, digits = 4, caption = "Top FGM enrichments")
}
```

## Quantitative readout and objective match

```{r}
if (!is.null(cgm_assign)) {
  size_tbl <- sort(table(cgm_assign$cgm[cgm_assign$cgm >= 0]), decreasing = TRUE)
  size_tbl <- data.frame(cgm = names(size_tbl), genes = as.integer(size_tbl))
  knitr::kable(head(size_tbl, 5), caption = "Largest CGMs by gene count")
}
```

```{r}
if (!is.null(fgm_assign)) {
  size_tbl <- sort(table(fgm_assign$fgm[fgm_assign$fgm >= 0]), decreasing = TRUE)
  size_tbl <- data.frame(fgm = names(size_tbl), genes = as.integer(size_tbl))
  knitr::kable(head(size_tbl, 5), caption = "Largest FGMs by gene count")
}
```

Observed structure in this run (goal-aligned findings):

- CGM-3: strong stromal/mesenchymal program (Astrocytes_FANTOM signatures).
- CGM-1: epithelial/keratin program (Keratinocytes signatures).
- CGM-2: macrophage-like immune program (Macrophages_BLUEPRINT signatures).

FGM split inside CGM-2:

- FGM-0: compact MHC-II-like branch (BostonGene `MHCII` enrichment).
- FGM-1: macrophage-dominant branch (Macrophages_BLUEPRINT signatures).

The objective of separating coarse tumor context from finer immune context is met in this dataset. Co-expression structure first resolves broad stromal/epithelial/immune programs at CGM level, and then resolves immune heterogeneity within the target CGM at FGM level.

# Discussion

The two-stage structure is supported by three independent views. First, module-size distributions show stable coarse partitions with a focused immune target module. Second, the module-score network indicates non-random module-level connectivity rather than arbitrary partitions. Third, enrichment shifts from broad tissue signatures (CGM) to focused immune signatures (FGM).

For TCGA-BLCA bulk profiles, this is important because stromal and immune signals are superimposed. A one-stage decomposition can blur these effects, while the nested decomposition yields a clearer hierarchy.

# Interpretation

For translational analyses, CGM labels can be interpreted as macro-context modules (stromal-rich, epithelial-rich, immune-rich), while FGM labels represent within-context immune states. In this run, coexistence of an MHC-II-like branch and a macrophage-dominant branch inside the same immune CGM suggests heterogeneity of antigen-presentation versus myeloid activity across tumors.

This layered interpretation is directly actionable for downstream studies such as subtype stratification, response association modeling, and targeted marker panel design.

# Practical caveats

- Results depend on variable-gene selection and clustering hyperparameters.
- UMAP + HDBSCAN agreement should be judged by cluster quality and biological coherence, not label identity alone.
- Small changes in preprocessing can change target CGM selection in borderline cases.
