---
title: "Tutorials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Background

In nested co-expression workflows, the first stage captures broad gene
organization and the second stage refines local structure after normalization.

## Objective

Run an end-to-end two-stage analysis without downloading any external data,
while preserving a realistic analysis pattern.

## Data

```{r}
library(NestedWGCNA)
set.seed(300)
n_samples <- 80
n_genes <- 600
x <- matrix(rnorm(n_samples * n_genes, sd = 0.8), nrow = n_samples, ncol = n_genes)
rownames(x) <- paste0("S", seq_len(n_samples))
colnames(x) <- paste0("G", seq_len(n_genes))
```

## Analysis

```{r}
res <- run_nested_wgcna(
  x = x,
  mode = "paper",
  min_cgm_size = 60,
  min_fgm_size = 20,
  top_n_genes = 350
)

s <- summary(res)
s
```

```{r}
table(res$cgm$assignment$module)
table(res$fgm$assignment$module)
dim(res$module_scores)
```

## Results

- CGM and FGM assignments are both produced from the same input matrix.
- Core genes are selected per CGM and stored in `res$cgm_core`.
- Module scores are returned as a compact sample-by-module matrix.

## Discussion

This minimal implementation prioritizes reproducible computation and low package
weight. It does not include heavy visualization or enrichment modules.

## Interpretation

The output object is designed so that downstream analyses can consume module
assignments and module scores directly without carrying large raw artifacts.
