---
title: "Tutorial: TCGA-BLCA real-data analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: TCGA-BLCA real-data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(NestedWGCNA)

locate_case_file <- function(name) {
  p <- file.path("inst", "extdata", "case_studies", "tcga_blca", name)
  if (!file.exists(p)) {
    p <- system.file("extdata", "case_studies", "tcga_blca", name, package = "NestedWGCNA")
  }
  p
}

summary_path <- locate_case_file("summary.json")
cgm_enr_path <- locate_case_file("cgm_enrichment.tsv")
fgm_enr_path <- locate_case_file("fgm_enrichment.tsv")

summary_obj <- if (file.exists(summary_path)) jsonlite::fromJSON(summary_path) else NULL
cgm_enr <- if (file.exists(cgm_enr_path)) read.delim(cgm_enr_path, check.names = FALSE) else NULL
fgm_enr <- if (file.exists(fgm_enr_path)) read.delim(fgm_enr_path, check.names = FALSE) else NULL
val_or_na <- function(x) if (is.null(x)) NA else x
```

# Background

Bulk RNA-seq co-expression analysis often mixes strong tissue-level signals with finer immune-state signals. A single module scale can miss this hierarchy. NestedWGCNA addresses this by first identifying broad modules (CGMs), then re-normalizing and re-clustering inside each module to reveal FGM structure.

# Objective

The objective of this tutorial is to run a fully reproducible, real-data NestedWGCNA workflow on TCGA-BLCA and interpret whether the two-stage design separates broad tumor programs from finer immune programs.

# Data source and download

This tutorial uses an openly available TCGA-BLCA expression matrix:

- URL: `https://raw.githubusercontent.com/ilyada/NestedWGCNA/main/data/TCGA_BLCA.tsv`
- Samples: 406
- Genes: 20,017

The analysis script downloads the expression matrix and signature references, runs the full R pipeline, and writes outputs under `inst/extdata/case_studies/tcga_blca/`.

```{r, eval = FALSE}
system("Rscript inst/scripts/run_tcga_blca_case_study.R")
```

# Methods

The executed case-study configuration is:

1. Log-transform expression and keep top 1,200 variable genes.
2. Run CGM discovery (`min_cluster_size = 50`) with `find_cgm()`.
3. Select an immune-like CGM by signature enrichment.
4. Attempt GenFocus (`focus = "eigengene"`, `corr_thr` ladder from 0.9 to 0.7, `CVR_thr = 0.6`).
5. If no INGS is established, run FGM discovery directly on the target CGM expression matrix.
6. Run FGM discovery in that CGM (`min_cluster_size = 10`).
7. Perform hypergeometric enrichment against xCell and BostonGene signatures.

```{r}
if (!is.null(summary_obj)) {
  method_tbl <- data.frame(
    item = c(
      "samples",
      "input genes",
      "modeled genes",
      "CGM min_cluster_size",
      "FGM min_cluster_size",
      "GenFocus corr_thr",
      "GenFocus INGS genes"
    ),
    value = c(
      summary_obj$n_samples,
      summary_obj$n_genes_input,
      summary_obj$n_genes_model,
      summary_obj$cgm_min_cluster_size,
      summary_obj$fgm_min_cluster_size_used,
      val_or_na(summary_obj$genfocus_corr_thr),
      summary_obj$genfocus_ings_n
    )
  )
  knitr::kable(method_tbl, caption = "Case-study analysis settings")
}
```

# Results

## Module recovery

```{r}
if (!is.null(summary_obj)) {
  res_tbl <- data.frame(
    layer = c("CGM", "FGM (inside target CGM)"),
    modules = c(summary_obj$cgm_n_modules, summary_obj$fgm_n_modules),
    assigned_genes = c(summary_obj$cgm_assigned_genes, summary_obj$fgm_assigned_genes),
    noise_genes = c(summary_obj$cgm_noise_genes, summary_obj$fgm_noise_genes)
  )
  knitr::kable(res_tbl, caption = "Recovered modules")
}
```

```{r, echo = FALSE, fig.cap = "CGM sizes"}
img <- "figures/tcga_blca_cgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

```{r, echo = FALSE, fig.cap = "FGM sizes in target CGM"}
img <- "figures/tcga_blca_fgm_module_sizes.png"
if (file.exists(img)) knitr::include_graphics(img)
```

## Biological enrichment

```{r}
if (!is.null(cgm_enr) && nrow(cgm_enr) > 0) {
  cgm_top <- cgm_enr[order(cgm_enr$fdr), c("module", "source", "signature", "k", "K", "fdr")]
  cgm_top <- head(cgm_top, 10)
  knitr::kable(cgm_top, digits = 4, caption = "Top CGM enrichments")
}
```

```{r}
if (!is.null(fgm_enr) && nrow(fgm_enr) > 0) {
  fgm_top <- fgm_enr[order(fgm_enr$fdr), c("fgm", "source", "signature", "k", "K", "fdr")]
  fgm_top <- head(fgm_top, 10)
  knitr::kable(fgm_top, digits = 4, caption = "Top FGM enrichments")
}
```

Observed structure in this run:

- CGM-3: strong stromal/mesenchymal program (Astrocytes_FANTOM signatures).
- CGM-1: epithelial/keratin program (Keratinocytes signatures).
- CGM-2: macrophage-like immune program (Macrophages_BLUEPRINT signatures).

FGM split inside CGM-2:

- FGM-0: compact MHC-II-like branch (BostonGene `MHCII` enrichment).
- FGM-1: macrophage-dominant branch (Macrophages_BLUEPRINT signatures).

# Discussion

The two-stage structure is supported by the output:

- Stage 1 recovered broad, high-level tissue programs.
- Stage 2 split the immune-like CGM into a small antigen-presentation branch and a larger macrophage branch.

This is exactly the expected use case for nested co-expression analysis in bulk tumors, where immune and stromal signals are often superimposed.

# Interpretation

For translational analyses, these outputs can be interpreted as:

- CGM labels: macro-context (stromal vs epithelial vs immune-rich).
- FGM labels: within-context immune states that may align with antigen presentation or macrophage activity.

In this run, coexistence of an MHC-II-like FGM and a macrophage-like FGM within the same immune CGM suggests heterogeneous antigen-presentation and myeloid states across BLCA tumors.

# Practical caveats

- Results depend on variable-gene selection and clustering hyperparameters.
- UMAP + HDBSCAN agreement should be judged by cluster quality and biological coherence, not label identity alone.
- Small changes in preprocessing can change target CGM selection in borderline cases.
